A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-07-03 text/manuscript.Rmd
reflow text-wrap paragraphs as lines for increased efficiency of git diffs

## Abstract
Cancer is a disease of cellular metabolism and other basic biochemical processes common to all cells. However, molecular signatures remain largely indistinguishable from tissue-of-origin, and general, unifying principles remain obscure. To address this challenge, we reduced mutation data to disruptions in a select set of pathways pointing to basic cell biology, from DNA replication to cellular communication. Using dimension reduction techniques, we assigned tumor samples into twelve classes distinct from their tissues-of-origin and largely free of bias from mutational burden or clinical stage. We showed that these tumor classes vary in prognosis by modeling relative risk of death in different classes and cancer types. We identified class-specific mutations in different tissues, demonstrating how tissue-specific signatures contribute to common cellular phenotypes. Moreover germline risk mutations involved in genome stability, such as BRCA1 and Rb1, are equally distributed among our classes, somewhat contrary to the expectation that such genes are hallmarks of molecular sub-types. We also investigated pathways that distinguish between metastatic and non-metastatic tumors, and found that most meaningful differences are class-specific. Most metastatic enriched pathways from one class in metastatic tumors are class-specific pathways from another class, suggesting phenotypic convergence, and potentially highlighting key pathways to focus on therapeutically for prevention of class-specific relapse. In summary, our observations suggest that some genes act as drivers of genome instability in all classes, while different sets of class-specific genes confer increased fitness in combinatorial fashion, eventually converging on one or more end-stage disease phenotypes. Thus, we present a coherent picture of global tumor biology, and explain how common cellular dysfunction might arise from tissue-specific mutations.

@@

#### Independence of Stage, mutation count and mutation profile.

@@

We find that most tumors segregate into one of 12 biologically distinct clusters, regardless of the cancer type (**figure 1h** and **supplementary figure S1**). Some cancer types ( _e.g._ pancreatic adenocarcinomas in the RAS arm clusters 9 and 10, and breast cancers being excluded from the same cluster, see **supplementary figure S1**) are unevenly distributed toward a specific class, we could not identify cancer type that was exclusive to a single tumor class. Somewhat surprisingly to us, this finding extends to histological subtypes of breast, head and neck cancers, leukemias, _etc._ (see **Figure 1i** and **Supplemental Figure S2**). Indeed, it would be surprising if we did not observe bias for some cancers, as it has long been known that certain treatment regimens are likely to work for a majority of patients of a given cancer type. Nonetheless, the basis for some tumors being treatment-refractory in spite of clinical indications remains elusive. Our analysis suggests the possibility of more general mechanisms underlying tumor evolution regardless of tissue, histological or cellular sub-type and could form a basis for future separation of patients into different precision treatment regimens.

>>>>

## Abstract
Cancer is a disease of cellular metabolism and other basic biochemical processes
common to all cells. However, molecular signatures remain largely
indistinguishable from tissue-of-origin, and general, unifying principles remain
obscure. To address this challenge, we reduced mutation data to disruptions in a
select set of pathways pointing to basic cell biology, from DNA replication to
cellular communication. Using dimension reduction techniques, we assigned tumor
samples into twelve classes distinct from their tissues-of-origin and largely
free of bias from mutational burden or clinical stage. We showed that these
tumor classes vary in prognosis by modeling relative risk of death in different
classes and cancer types. We identified class-specific mutations in different
tissues, demonstrating how tissue-specific signatures contribute to common
cellular phenotypes. Moreover, germline risk mutations involved in genome
stability, such as BRCA1 and Rb1, are equally distributed among our classes,
somewhat contrary to the expectation that such genes are hallmarks of molecular
subtypes. We also investigated pathways that distinguish between metastatic and
non-metastatic tumors, and found that most meaningful differences are
class-specific. Most metastatic enriched pathways from one class in metastatic
tumors are class-specific pathways from another class, suggesting phenotypic
convergence, and potentially highlighting key pathways to focus on
therapeutically for prevention of class-specific relapse. In summary, our
observations suggest that some genes act as drivers of genome instability in all
classes, while different sets of class-specific genes confer increased fitness
in a combinatorial fashion, eventually converging on one or more end-stage
disease phenotypes. Thus, we present a coherent picture of global tumor biology,
and explain how common cellular dysfunction might arise from tissue-specific
mutations.

@@

#### Independence of stage, mutation count and mutation profile.

@@

We find that most tumors segregate into one of 12 biologically distinct
clusters, regardless of the cancer type (**figure 1h** and **supplementary
figure S1**). Some cancer types ( _e.g._ pancreatic adenocarcinomas in the RAS
arm clusters 9 and 10, and breast cancers being excluded from the same cluster,
see **supplementary figure S1**) are unevenly distributed toward a specific
class, we could not identify cancer type that was exclusive to a single tumor
class. Somewhat surprisingly to us, this finding extends to histological
subtypes of breast, head and neck cancers, leukemias, _etc._ (see **Figure 1i**
and **Supplemental Figure S2**). Indeed, it would be surprising if we did not
observe bias for some cancers, as it has long been known that certain treatment
regimens are likely to work for a majority of patients of a given cancer type.
Nonetheless, the basis for some tumors being treatment-refractory in spite of
clinical indications remains elusive. Our analysis suggests the possibility of
more general mechanisms underlying tumor evolution regardless of tissue,
histological or cellular subtype and could form a basis for future separation of
patients into different precision treatment regimens.
